On the Functionality of the Multilayer Perceptron

### 2020-12-09 Thesis/Drafts/Buell-PHYS799-ThesisDraft0.tex
Updates to properties section

\paragraph*{}As discussed in Sec. (\ref{sec-TheNeuralNetwork}), performance quality of a neural network is greatly dependent on the properties of the chosen features \cite{Virtanen,Liu}. To ensure that a the classifier model performs adequately and consistently, we must choose these features to have high variance between classes, and low variance within each class. This enables a neural network to development clear decision boundaries between each unique class \cite{James,Serizel}. To develop this set of features, we explore the physical and mechanical properties of musical instruments and the sounds that they create.

\paragraph*{}We have assembled training data samples than can be grouping into $37$ classes of unique sources listed below:

@@

\paragraph*{}Rather than explore each class individually, we can examine the properties of groupings of musical instruments by using the \textit{Hornbostel-Sachs} organization system. In this syetem, musical instruments or sound sources can be divided up into four broad categories based on the nature of the sound-producing material. These categories are (i) idiophones, (ii) membranophones, (iii) chordophones, and (iv) aerophones. In this section, we explore the mechanical properties of each group and show how the physics of the source instrument influences our feature choice.

@@

\paragraph*{}An idiophone is an instrument that produces sound through the vibration of the full body of the object. This generally includes most percussive instruments excluding drums \textcolor{red}{citation needed}. From the set of classes that we use, bells, crotales, Hi-hats, vibraphones, and xylophones are all examples of idiophones.

@@

@@

\paragraph*{}A chordophone is a musical instrument that produces sound through the vibration of strings that are stretched between two fixed points \textcolor{red}{citation needed}. From our data set, banjos, basses, cellos, guitars, mandolins, violins, and violas are all examples of chordophones.

\subsubsection{The 1-Dimensional Vibrating String}

\paragraph*{}We can model a chordophone as having a string of length $L$, with a constant linear mass density, $\mu$, subject to a tension force $F_T$. The string is flexible, and free to vibrate in a single dimension. We describe the state of string over space and time with a function $U(x,t)$. The behavior of the string can be modeled by D'Alembert's 1-dimensional wave equation \cite{Haberman,Hunter,Taylor}:
\begin{equation}
\label{eqn-1DWaveEqn}
\frac{\partial^2 U}{\partial t^2} = v^2 \frac{\partial^2 U}{\partial x^2}
\end{equation}
Where $v$ is the wave propagation velocity, given by $\sqrt{\frac{F_T}{\mu}}$. D'Alembert's equation indicates that the acceleration of any point on a string is directly proportional to the rate of change of the slope of the string at that point. Additionally, the string is fixed in place at both ends, with a nut at $x = 0$ and a bridge at $x = L$, which can be modeled by the boundary conditions \cite{Olson,White,Haberman}:
\begin{equation}
U(0,t) = 0 
	\text{\hspace*{2cm} and \hspace*{2cm}}
U(L,t) = 0 
\end{equation}

\paragraph*{}In order to specify a solution, we must also specify two initial conditions. We described the initial state of the string at time $t = 0$  by the two conditions:
\begin{equation}
U(x,0) = f(x) 
	\text{\hspace*{2cm} and \hspace*{2cm}}
\frac{\partial U}{\partial t}(x,0) = g(x)
\end{equation}
Depending on the exact nature of the source instrument, $f(x)$ and $g(x)$ can take on different forms. For example, for plucked stringed instruments, $f(x)$ may appear to be as triangle-like function, with a peak at some position $0 < x_0 < L$ where then string was struck. It is these initial conditions that allow for the drastically different timbres that arise in the wave forms.

\paragraph*{}Despite these differences, the wave equation is a \textit{linear} partial differential equation which allows for the fact that a linear combination of solutions is also a valid solution. In the case of the taught finite string that is fixed at both ends, such a guitar, viola or cello, the a solution tends to take the form of a sinusoidal function such a sine or cosine \cite{Hunter, Taylor, White}.

\paragraph*{}The boundary continuations, and most forms of standard initial conditions allow for the each sinusoidal to fit exactly an integer wavelengths into the region $[0,L]$ \cite{Haberman,Hunter}. Physically, this corresponds to having the string vibrate with some fundamental frequency $f_0$, which represents it's lowest possible note,and a linear combination of higher harmonics, $n \times f_0$, where $n \in \mathbb{Z}$ \cite{Olson,Taylor,White}.

\begin{figure}[H]
	\begin{subfigure}{0.45\textwidth}
	\centering
	\includegraphics[scale=0.18]{../FiguresSpectrogram/CELLO-F4}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
	\centering
	\includegraphics[scale=0.18]{../FiguresSpectrogram/Guitar-B2}
	\end{subfigure}
\end{figure}

@@

\paragraph*{}An aerophone is a musical instrument that produces sounds through the vibration of an air molecules in a column-like structure.

@@

\paragraph*{}We group sounds produced from non-physical musical instruments in this final category. Form our data set, this category includes the four simplest waveforms, (i) sine waves, (ii) sawtooth waves, (iii) square waves, and (iv) triangle waves, as well as any color noises \cite{White}. Since these sounds are only produced synthetically through a MATLAB program, we choose not to include them in the Hornbostel-Sachs grouping above. We explore the properties of each of these generated sounds.

\subsubsection{Sine Wave}

\paragraph*{}A sine wave is the simplest of sound waves, and makes up the foundation for all periodic wave forms \cite{White,Virtanen}. It consists of a single oscillatory term with a fixed fundamental frequency $f_0$. A sine wave's amplitude as a function of time is given by 
\begin{equation}
\label{eqn-SineWave}
x(t) = A_0 \sin(2\pi f_0 t + \phi)
\end{equation}
Where $A_0$ gives an amplitude, $f_0$ is the fundamental frequency, and $\phi$ is a phase shift. We show a simple sine wave in the time and frequency domain in Fig. (\ref{fig-SineWave}).

\begin{figure}[h]
\begin{center}
	\textit{[Image of Sine Wave Waveform]}
	\textit{[Image of Sine Wave Freq. Spect]}
\end{center}
\caption{A sine wave in the (a) time-domian and (b) frequency domain}
\label{fig-SineWave}
\end{figure}

\paragraph*{}A pure sine wave contains no overtones, only a time-independent fundamental frequency, which can take on any value due to synthetic generation.

\subsubsection{Sawtooth Wave}

\subsubsection{Square Wave}

\paragraph*{}A square wave is a piece-wise defined waveform whose value alternatively alternates between $+1$ and $-1$ \cite{White,Olson}. A square wave with period $f_0$ has a waveform defined over a single period as:
\begin{equation}
\label{eqn-SquareWave}
x(t) =  \left\{
\begin{array}{ll}
    +1 & \text{\hspace{0.5cm}} 0 < x \leq T_0 / 2 \\
    -1 & \text{\hspace{0.5cm}} T_0/2 < x < T_0
\end{array} 
\right.
\end{equation}
where $T_0 \equiv 1/f_0$.

\paragraph*{}We can use Fourier decomposition to examine the constituent sinusoidal wave that make up a square wave. The frequency space representation of a square wave indicates that is can be approximated as a sum of higher odd-numbered harmonics, with the amplitude of each decreasing by the reciprocal of the harmonic number \cite{White}. Thus a square wave in the time domain can be approximated as:
\begin{equation}
x(t) \approx \sum_{n = 1}^{\infty} \frac{1}{2n-1} \sin \bigg( 2 \pi (2n - 1)f_0 t \bigg)
\end{equation}

\begin{figure}[h]
\begin{center}
	\textit{[Image of Square Wave Waveform]}
	\textit{[Image of Sqaure Wave Freq. Spect]}
\end{center}
\caption{A square wave in the (a) time-domain and (b) frequency domain}
\label{fig-SquareWave}
\end{figure}

\subsubsection{Triangle Wave}

\subsubsection{White Noise}

@@

>>>>

\paragraph*{}As discussed in Sec. (\ref{sec-TheNeuralNetwork}), performance quality of a neural network is greatly dependent on the properties of the chosen features \cite{Virtanen,Liu}. To ensure that the classifier model performs adequately and consistently, we must choose these features to have high variance between classes, and low variance within each class. This enables a neural network to development clear decision boundaries between each unique class \cite{James,Serizel}. To develop this set of features, we explore the physical and mechanical properties of musical instruments and the sounds that they create.

\paragraph*{}Computer algorithms such are neural networks are build to handle exclusively quantitative information \cite{Geron}. Where a human would use qualitative adjectives such as \textit{bright} or \textit{harsh} or \textit{percussive} to describe sound, computers require a numerical descriptor instead. The goal of this section is to develop the attributes of sound that enable us to describe a sound wave in a quantitative and compact way. Thus we use values that can be expressed as numbers such as \textit{amplitude envelopes}, \textit{frequency spectra}, or \textit{formant structure}. For this section, we represent waveforms as \textit{spectrograms} to capture the time-space and frequency-space development of energy simultaneously. We motivate and describe the production of a spectrogram in sec. (\ref{subsec-Spectrogram}).

\paragraph*{}We have assembled training data samples than can be grouped into $37$ classes of unique sources listed below:

@@

\paragraph*{}Rather than explore each class individually, we can examine the properties of groupings of musical instruments by using the \textit{Hornbostel-Sachs} organization system developed by  Erich M. von Hornbostel and Curt Sachs, and published in 1961 \cite{Hornbostel}. In this system, musical instruments or sound sources can be divided up into four broad categories based on the nature of the sound-producing material. These categories are (i) idiophones, (ii) membranophones, (iii) chordophones, and (iv) aerophones. In this section, we explore the mechanical properties of each group and show how the physics of the source instrument influences our feature choice.

@@

\paragraph*{}An idiophone is an instrument that produces sound through the vibration of the full body of the object. This generally includes most percussive instruments excluding drums \cite{Hornbostel}. From the set of classes that we use, bells, crotales, Hi-hats, marimbas, vibraphones, and xylophones are all examples of idiophones. In each case, sound comes from the vibration of the object itself or some subset of the body. In the case of mallet percussion, the wood or metal keys vibrate when struck with a felt or yarn mallet. Sometimes the vibration is amplified by nearby air columns, but it is the vibration of the key itself where the sound originates from \cite{Olson}.

\paragraph*{}Idiophones in particular can be characterized by their \textit{transient response} \cite{White}. The transient response is the time-evolution response of the waveform in time, which we normally break into \textit{attack}, \textit{decay}, \textit{sustain}, and \textit{release}. For most idiophones, the system is struck once and left to a free or dampened vibration; which we can mathematically model as a rigid body reacting to an impulse \cite{Hunter,Olsen}.

\begin{figure}
	\begin{subfigure}[b]{0.45\textwidth}
	\centering
	\textcolor{red}{Idiophone}
	\caption{•}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
	\centering
	\textcolor{red}{Idiophone}
	\caption{•}
	\end{subfigure}
\caption{Spectrogram representation of idiophone waveforms}
\label{fig-PropertiesIdiophones}
\end{figure}

\paragraph*{}From the spectrogram representation of idophone waveforms , we can see the majority of energy in the time domain is right at the start of the waveform. This is shown by the vertical bright yellow line at on the left side of each image in Fig. (\ref{fig-PropertiesIdiophones}). When an idiophone is struck, all of the energy for the vibration is delivered up front, and this the amplitude of the waveform can only decay. This gives the resulting transient a very short attack time, followed by a short decay, sustain, and often long release as the last of the energy dies out \cite{White}. While other instruments sch as plucked strings like guitars show similar response, this is a very important property

\paragraph*{}This property allows for us to distinguish idiophones from other instruments by an approximation of their amplitude envelope. For example, if we divide the time series waveform into $q$ distinct sections, and compute the RMS energy in each section, we should see a gradual decay of energy with time. More specifically, the energy in some section $q_i$ should be greater than the energy in any subsequent section $q_{i + n}, n \in \mathbb{Z}$. Having this property does not guarantee the source is an idiophone but can we can reasonable state that most idiophones follow this behavior. In addition to the amplitude envelope, we can consider the distribution of time-space energy in the form of a single scalar quantity as the center of mass. If we treat the waveform as a 1D mass distribution, and compute the center of mass, we should see consistently low values because the largest values of amplitudes happen early on in the time-domain.

@@

\paragraph*{}A membranophone is an instrument that produces sound through the vibration of a tight membrane \cite{Hornbostel}. This includes most drum-like instruments While none of the instruments that we classify are membranophones, they do exhibit some noteworthy properties that influence our feature for other instruments. Common examples include the tympani, the bass drum, or the kettle drum. Each of these instruments is composed of a large cylinder or bowl-like structure that serves as an amplifying cavity \cite{Olson,White}. Given that most membranophone instruments are played by striking the membrane with the hand or mallet, the amplitude envelope properties

\paragraph*{}Additionally, we model the tight elastic membrane very similarly as that of a tight elastic string, except in two dimensions \cite{White}. However, the partials for the membrane are not related by integer multiples like that or strings are. This lack of harmonic structure leaves most membranophones to have a sort of "muddled" frequency spectrum where we see a continuous distribution of constituent frequencies arise as opposed to clean overtones arising as in stringed or wind instruments \cite{White,Olson}. This is in part why many membranophones have a much more "percussive" sound as opposed to performing any distinct note.

@@

\paragraph*{}A chordophone is a musical instrument that produces sound through the vibration of strings that are stretched between two fixed points \cite{Hornbostel}. From our data set, banjos, basses, cellos, guitars, mandolins, violins, and violas are all examples of chordophones. We commonly model a chordophone as having a string of length $L$, with a constant linear mass density, $\mu$, subject to a tension force $F_T$. The string is flexible, and free to vibrate in a single dimension. We describe the state of string over space and time with a function $U(x,t)$. The behavior of the string can be modeled by D'Alembert's 1-dimensional wave equation \cite{Haberman,Hunter,Taylor}:
\begin{equation}
\label{eqn-1DWaveEqn}
\frac{\partial^2 U}{\partial t^2} = v^2 \frac{\partial^2 U}{\partial x^2}
\end{equation}
Where $v$ is the wave propagation velocity, given by $\sqrt{\frac{F_T}{\mu}}$. D'Alembert's equation indicates that the acceleration of any point on a string is directly proportional to the rate of change of the slope of the string at that point.

\paragraph*{}Since the PDE is second order in time, we must also specify two initial conditions being the shape and the derivative of the shape at time $t = 0$.  \cite{Olson,White,Haberman}:
\begin{equation}
U(x,0) = f(x) 
	\text{\hspace*{2cm} and \hspace*{2cm}}
\frac{\partial U}{\partial t}(x,0) = g(x)
\end{equation}
Additionally, the string is fixed in place at both ends, with a nut at $x = 0$ and a bridge at $x = L$, which can be modeled by the boundary conditions
\begin{equation}
U(0,t) = 0 
	\text{\hspace*{2cm} and \hspace*{2cm}}
U(L,t) = 0 
\end{equation}
Depending on the exact nature of the source instrument, $f(x)$ and $g(x)$ can take on different forms. For example, for plucked stringed instruments, $f(x)$ may appear to be as triangle-like function, with a peak at some position $0 < x_0 < L$ where then string was struck. It is these initial conditions that allow for the drastically different timbres that arise in the wave forms. Compare the spectrogram representation of a few stringed instruments in Fig. (\ref{fig-PropertiesChordophones}).
\begin{figure}[H]
	\begin{subfigure}{0.45\textwidth}
	\centering
	\includegraphics[scale=0.18]{../FiguresSpectrogram/CELLO-F4}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
	\centering
	\includegraphics[scale=0.18]{../FiguresSpectrogram/Guitar-B2}
	\end{subfigure}
\caption{Spectrograms for a few Stringed instruments}	
\label{fig-PropertiesChordophones}
\end{figure}

\paragraph*{}Despite these differences, the wave equation is still a \textit{linear} partial differential equation which allows for the fact that a linear combination of solutions is also a valid solution. In the case of the taught finite string that is fixed at both ends, such a guitar, viola or cello, the a solution tends to take the form of a sinusoidal function such a sine or cosine \cite{Hunter, Taylor, White}. The boundary continuations, and most forms of standard initial conditions allow for the each sinusoidal to fit exactly an integer wavelengths into the region $0 \leq x \leq L$ \cite{Haberman,Hunter}. Physically, this corresponds to having the string vibrate with some fundamental frequency $f_0$, which represents it's lowest possible note,and a linear combination of higher harmonics, $n \times f_0$, where $n \in \mathbb{Z}$ \cite{Olson,Taylor,White}. In the case of bowed stringed instruments such as cellos or violins, we often see a formant structure marked by fairly uniform amplitude of each harmonic in the frequency spectrum. Observe this behavior in Fig. (\ref{fig-PropertiesChordophones}).

\paragraph*{}Additionally, It is worth noting at each string instrument produces a slightly different frequency spectrum, hence each provides a slightly different timbre to the listener, so differentiating between them is difficult, but possible. Because of this, we find that the spectrogram representation of the waveform is so critical to classification success because the convolution layers allow for the detection of small variations in the frequency-time domain where humans would otherwise be unable to \cite{Loy,Virtanen}. However, the size of the instrument limits the length of the string that vibrates which leads to limitations in the range of the overtones the instruments will produce. For example, a cello playing a low $C1$ will have overtones extenting up into the $12$ kHz range, but the energy will be very low there, thus have a very low frequency center of mass. Compare this with a violin playing a $C6$, will have overtones into the $20$ kHz range, and will produce a very high frequency center of mass. While the ranges of the musical instruments overlap a great deal, the energy distribution in frequency space can be treated as a 1D mass distribution, and the frequency center of masses of instruments with similar timbres can still differ greatly \cite{Serizel}.

\paragraph*{}Notice as well in the spectrograms of Fig. ({fig-PropertiesChordophones}), how the transient response (the time domain) of chordophones differs wildly from idiophones and membranophones. In the case of bowed instruments, the attack of the stringed instrument is much longer, with a much more gradual sustain decay, and release. Where an idiophone or plucked string has all of the time-domain energy up front, the bowed string waveform continues to build energy as the instrument is bowed, and only loses energy once the bowing stops, and the string is left to vibrate by itself \cite{Hunter}. This causes a bowed stringed instruemnt to present a very different amplitude envelope compared to the other categories of instruments, and a much more centrally located temporal center of mass.

@@

\paragraph*{}An aerophone is a musical instrument that produces sounds through the vibration of an air molecules in a column-like structure. This is the largest category of instruemnts that we have, and is commonly broken down into smaller subcategories such as \textit{woodwind} and \textit{bass} instruments. From out data set- alto flutes, alto saxophones, bass clarinets, bass flutes, bassoons, $B\flat$ clarinets, contrabassoons, $E\flat$ clarinets, English horns, flutes, oboes, tenor saxophones, and soprano saxophones are all examples of woodwind aerophones. Additionally, bass trombones, french horns, tenor trombones, trombones, $B\flat$ trumpets, and tubas are all examples of brass aerophones \cite{Hornbostel,Olson,White}.

@@

\paragraph*{}We group sounds produced from non-physical musical instruments in this final category. Form our data set, this category includes the four simplest waveforms,sine waves, (ii) sawtooth waves, square waves, and triangle waves, as well as any color noises \cite{White}. Since these sounds are only produced synthetically through a MATLAB program, we choose not to include them in the Hornbostel-Sachs grouping above. We explore the properties of each of these generated sounds and how they relate to physical instruments.

@@

\bibitem{Hornbostel}
Von Hornbostel, Erich, and Curt Sachs. “Classification of Musical Instruments.” \textit{The Galpin Society Journal}, Translated by Anthony Baines and Klaus Wachsmann, vol. 14, Mar. 1961, pp. 3–29.
